---
title: 'PSTAT 115: Bayesian Data Analysis'
author: "Professor Laura Baracaldo"
output:
  xaringan::moon_reader:
    css: ['default', "xaringan_style.css"]
    nature:
      highlightLines: true
      ratio: "4:3"
---


```{r setup, include=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
opts_chunk$set(echo=FALSE, fig.align='center', out.width="70%") 
ig <- function(file) {knitr::include_graphics(file)}
source("../scripts/make_figs.R")
source("../scripts/color_defs.R")
```


### Class Resources 
__Required Textbook__

- Bayes Rules: https://www.bayesrulesbook.com/

__Course Pages__

- Nectir for course related questions and discussion: https://app.nectir.io/us/ucsb/group/PSTAT115-Fall23

- Gradescope: https://www.gradescope.com/courses/640308


---

### Grades

- 35% - expect approximately 4 homeworks 

- 20% - Midterm (Nov 1st)

- 15% - Quizzes

- 30% - Final exam (Dec 6th)

---

### Homework

- There will be approximately 4 homeworks (35% of your grade total)

- You will typically have 2 weeks to complete the homework

- Homework turned in within 24 hrs after the deadline without prior approval will receive a 20% deduction

- Homework will not be accepted more than 24 hrs late.

---

### Homework submission format

- All code must be written to be reproducible in Rmarkdown

- All derivations can be done in any format of your choosing (latex, written by hand) but must be legible and _must be integrated into your Rmarkdown pdf_.

- Ask a TA _early_ if you have problems regarding submissions. 

---

### Software and Deliverables

__Software__

- [R](http://cran.r-project.org/) ([R studio](http://www.rstudio.com/))



__Homeworks submission format__

- Electronic submission via Gradescope

- [R markdown code](http://rmarkdown.rstudio.com/)

- Generated PDF file with eventual handwritten notes 

---

### Labs and Quizzes

- There will be a handful of quizzes throughout the quarter.  

- The quizzes will be on Canvas.  

- You will have 10 minutes to take the quiz any time within a window of 12 hours.

- The quizzes will be given on lecture days

- There are no makeups, but the lowest quiz grade will be dropped from your final score.

- Quizzes will be multiple choice and will test your comprehension of the basic concepts.


---

### Class Policies

- All questions should be posted on nectir, _not by email_ (unless they are personal or grade-related)

---

### RStudio Cloud Service


- Log on to [pstat115.lsit.ucsb.edu](https://pstat115.lsit.ucsb.edu)
    + Cloud based rstudio service
    + Log in with your UCSB NetID

- Use [Jupyter Hub](https://pstat115.lsit.ucsb.edu/hub/user-redirect/git-pull?repo=https%3A%2F%2Fgithub.com%2Flnbaracaldol%2FPSTAT115-F23.git&urlpath=tree%2FPSTAT115-F23.git%2F&branch=main) to sync new material (BOOKMARK THIS)
  
- Make sure you can write and compile an [R
  markdown](http://rmarkdown.rstudio.com) (`Rmd`) document online
  
- Text formatting is minimal but
  [syntax](http://rmarkdown.rstudio.com/authoring_pandoc_markdown.html) is
  simple
---

### Markdown and mathematical formulas
The text inserted between two `$` signs will be interpreted as a Latex instruction, e.g.
`$x$`

Code                                       | Rendered math 
-------------------------------------------|-------------------------------------------
`$x$`                                      | $x$
`$\theta$`                                 | $\theta$
`$x_i^2$`                                  | $x_i^2$
`$\frac{1}{n}\sum_{i=1}^n x_i$`            | $\frac{1}{n} \sum_{i=1}^n x_i$
`$\frac{1}{n}\sum_{i=1}^n (x_i-\bar x)^2$` | $\frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2$

Rmarkdown and Latex resources:
<li><a href="http://rmarkdown.rstudio.com/lesson-1.html" target="_blank">Introduction to RMarkdown</a></li>
<li><a href="https://wch.github.io/latexsheet/" target="_blank">Latex cheat sheet</a></li>
<li><a href="http://www.andy-roberts.net/writing/latex" target="_blank">Introduction to Latex</a></li>

---

### Other R resources

- Cheatsheets: [https://www.rstudio.com/resources/cheatsheets/](https://www.rstudio.com/resources/cheatsheets/)

- _An Introduction to R_ - Venables and Smith
  
    http://cran.r-project.org/doc/manuals/R-intro.pdf

- _Using R for Introductory Statistics_ - John Verzani

    http://cran.r-project.org/doc/contrib/Verzani-SimpleR.pdf

- _R Markdown reference_ - https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf

- Probability cheatsheet in resources folder of cloud environment
---
class: center, middle

# What is Bayesian statistics?
---
class: center, middle
# What is the version of statistics you already know?
---
### Frequentist statistics

<center> What you learned in PSTAT 120B </center>
<br>

- Associated with the _frequentist_ interpretation of probability

  + For any given event, only one of two possibilities may hold: it occurs or it does not.

  + The _frequency_ of an event (in repeated experiments) is the _probability_ of the event

---
### Frequentist probability
<center>The probability of a coin landing on heads is 50%</center>
```{r, echo=FALSE, fig.align='center', out.width='50%'}
n <- 1000
plot(cumsum(rbinom(n, p=0.5, size=1))/(1:n), ylim=c(0,1), type="l", xlab="Flips", ylab="Fraction of 'heads'", col="red", lwd=2, cex.lab=1.5)
abline(h="0.5")
```
<center>The long run fraction of heads is 50%</center>
---
### Frequentist statistics

<center> What you learned in PSTAT 120B </center>
<br>

- Null Hypothesis Significant Testing (NHST) and Confidence Intervals
    + Frequentist uncertainty premised on imaginary resampling of data
    
    + Example: If the null model is true, and I re-run the experiment many times, how often will I reject?

---

### Confidence intervals

I have a $95\%$ confidence interval for a parameter $\theta$.  What does this mean?
<!-- >- $Pr(l(Y) < \theta < u(Y)\mid \theta) = 0.95$ -->
<!-- >- $Y$ is a random so $l(Y)$ and $u(y)$ is as well! -->
<!-- >- $\theta$ is an unknown constant -->
<!-- + Example: If I re-run the experiment many times what fraction of times will the interval cover the true parameter? -->
```{r, out.width='55%'}
prob <- 0.6
nintervals <- 50
sz <- 20
counts <- rbinom(n=nintervals, size=sz, prob=prob)
phat <- counts / sz

lower <- phat - 1.96 * sqrt(phat * (1-phat) / sz)
upper <- phat + 1.96 * sqrt(phat * (1-phat) / sz)

plot(0, 0, ylim=c(-1, nintervals + 1), xlim=c(0, 1), cex=0, xlab="p", ylab="")
segments(x0=lower, x1=upper, y0=1:nintervals, y1=1:nintervals, col=ifelse(upper < prob | lower > prob, "red", "grey"), lwd=1.5)
abline(v=prob, col="blue")

```
We expect $0.05 \times 50 = 2.5$ of the intervals to _not_ cover the true parameter, $p = 0.6$, on average

---

### Hypothesis Testing

```{r, out.width="60%"}
ig("images/blackswan.jpg") ## example 1.2.2 in rethinking

## Europeans had only ever seen white, went to australia saw black
## We've rejected our null, learned something
## Most of the time, science is not so logically dicsrete

```
<br>
<br>
$H_0:$ "All swans are white" vs $H_A:$ "not all swans are white".

- What is the probability of rejecting the Null? -If we observe at least one back swan, we reject: Depends on how rare black swans are.
- A more complex null hypothesis: At least 50% of swans are black.

---

###  Win probability

```{r, out.width="60%"}
ig("images/win_probability.png")
```
<center>source: fivethirtyeight.com</center>

---
###  Win probability

```{r, out.width="70%"}
ig("images/election_win_prob.png")
```
<center>source: fivethirtyeight.com</center>

---

### Bayesian probability

```{r, out.height= "400px", out.width="300px"}
ig("images/theory_of_probability.jpg")
```

<center>Bruno de Finetti begean his book on probability with:</center>
<center>"PROBABILITY DOES NOT EXIST"</center>
---
### Bayesian probability

- de Finetti is arguing that probability is about _belief_

  + Probability doesn't exist in an _objective_ sense

  + "The coin is fair" means _I believe_ that its equally likely to be heads or tails.

  + "Hillary Clinton has a 71% chance to win" reflects a belief, since the election happens only once

- Rarely, if ever, get _true_ replications to estimate frequentist probabilities

- Bayesian idea: focus statistical practice around belief about parameters

---
### Bayesian probability

<br>
<br>
<br>
<br>

> "The terms _certain_ and _probable_ describe the various degrees 
> of rational belief about a proposition which different amounts 
> of knowledge authorise us to entertain. All propositions are true 
> or false, but the knowledge we have of them depends on our circumstances
>
> `r tufte::quote_footer('--- John M Keynes')`

---

### Perceptions of Probability

```{r, out.width="70%"}
ig("images/perceptions.png")
```
source: https://github.com/zonination/perceptions
---

### Why Bayesian statistics?

- Classic statistical toolbox may not be appropriate for all settings.  

    + Inflexible and fragile
    + e.g. what if the assumptions of the test don't hold?

- Bayesian statistics provides a procedure for building our own tests / tools.

    + Design, build and refine procedures for you own models.

- A variety of powerful tools for inference with computer simulation

- Philosophy of science: quantifying degrees of belief often a more useful perspective than falsification

---

### Setup

- The _sample space_ $\mathcal{Y}$ is the set of all possible datasets.  
    + $Y$ is a random variable with support in $\mathcal{Y}$
    + We observe one dataset $y$ from which we hope to learn about the world.  

- The _parameter space_ $\Theta$ is the set of all possible parameter values $\theta$

- $\theta$ encodes the population characteristics that we want to learn about!

---

### Three steps of Bayesian data analysis

1. Construct a plausible probability model governed by parameters $\theta$

    + This includes specifying your belief about $\theta$ before seeing data (_the prior_)

2. Condition on the observed data and compute _the posterior_ distribution for $\theta$

3.  Evaluate the model fit, revise and extend.  Then repeat.

---
### Bayesian Inference in a Nutshell

1.  The _prior distribution_ $p(\theta)$ describes our belief about the true population characteristics, for each value of $\theta \in \Theta$.

2.  Our _sampling model_ $p(y\mid \theta)$ describes our belief about what data we are likely to observe if $\theta$ is true.  

3.  Once we actually observe data, $y$, we update our beliefs about $\theta$ by computing _the posterior distribution_ $p(\theta \mid y)$.  We do this with Bayes' rule!

<br>
&nbsp;&nbsp;&nbsp; **Key difference: $\theta$ is random!**

---
### Bayes' Rule

$$ P(A \mid B) = \frac{P(B \mid A)P(A)}{P(B)}$$

- $P(A \mid B)$ is the conditional probability of A given B
- $P(B \mid A)$ is the conditional probability of B given A
- $P(A)$ and $P(B)$ are called the marginal probability of A and B (unconditional)
---

### Bayes' Rule for Bayesian Statistics

$$ P(\theta \mid y) = \frac{P(y \mid \theta)P(\theta)}{P(y)}$$


- $P(\theta \mid y)$ is the posterior distribution

- $P(y \mid \theta)$ is the likelihood

- $P(\theta)$ is the prior distribution

- $P(y) = \int_\Theta p(y\mid \tilde \theta)p(\tilde \theta) d\tilde \theta$ is the model evidence 

---

### Bayes' Rule for Bayesian Statistics

$$P(\theta \mid y) = \frac{P(y \mid \theta)P(\theta)}{P(y)}\propto P(y \mid \theta)P(\theta)$$

- Start with a subjective belief (prior)

- Update it with evidence from data (likelihood)

- Summarize what you learn (posterior)

---

### Example: Estimating COVID Infection Rates 

- We need to estimate the prevalence of a COVID in Isla Vista

- Get a small random sample of 20 individuals to check for infection

```{r}
ig("images/covid.jpg")
```

---

### Example: Estimating Infection Rates 

- $\theta$ represents the population fraction of infected

- $Y$ is a random variable reflecting the number of infected in the sample

- $\Theta = [0,1] \quad \mathcal{Y} = \{0,1 , \ldots , 20 \}$

- Sampling model: $Y \sim \text{Binom}(20, \theta)$

---
### Example: Estimating Infection Rates 

```{r, out.width="50%"}
par(cex = 1.5)
fig_1_1_a(cols5[1:3])
```

---

### Example: Estimating Infection Rates 

- Assume _a priori_ that the population rate is low 
        
    + The infection rate in comparable cities ranges from about 0.05 to 0.20
    
- Assume we observe $Y = 0$ infected in our sample

- What is our estimate of the true population fraction of infected individuals?

---

## Example: Estimating Infection Rates 

```{r, out.width="50%"}
par(cex = 1.5)
fig_1_1_b(cols2)
```

---
### Tentative syllabus

- One parameter models (binomial, poisson, and normal)

- Monte Carlo methods (i.e. simulation-based inference)

- Markov chain Monte Carlo (MCMC)

- Hierarchical modeling

- An introduction to probabilistic programming

---
### Assignment

- Check Nectir, bookmark important links

- Start reviewing probability cheat sheet!

- Read chapters 1 and 2 of Bayes Rules